Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2015-04-06 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Update figure names to remove ., update latex and bib file

A free standing steel frame wall with a 12.7~mm thick, 2.44~m high by 1.22~m wide, calcium silicate board face was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux. Marinite Type I is the commercial name of the non-combustible board used in these experiments.  It is normally used as a fire-resistant thermal insulation that also has structural strength. 90~\% of the board is composed of calcium silicate and calcium meta-silicate.  The remaining 10~\% of the board is composed of a combination of organic fibers, fiber glass filament and crystalline silica (quartz)~\cite{BNZ:2012}.  Material properties that impact heat transfer were selected from the manufacturer's literature and are presented in the table below~\cite{Marinite:1997}.

\begin{table}
	\centering
\begin{tabular}{|c|c|c|c|c|}
	\hline Density & \multicolumn{2}{c|}{Thermal Conductivity} & \multicolumn{2}{c|}{Specific Heat}   \\
    \hline (kg/m$^3$) & Temperature ($^\circ$C) & (W/mK)  & Temperature ($^\circ$C)  & (kJ/kg $^\circ$C) \\
	\hline 737  & 24 & 0.13 & 93 & 0.28 \\
	\hline  & 205 & 0.12  & 205 & 0.30 \\
	\hline  & 316 & 0.11 & 316 & 0.32 \\
	\hline  & 425 & 0.12 & 425 & 0.34 \\
	\hline  & 538 & 0.12 &  &  \\
	\hline
\end{tabular}
\caption[Thermal properties for Marinite I]{Thermal properties for Marinite I, as reported by the manufacturer BNZ~\cite{Marinite:1997}
\label{tab:Marinite_Thermal_Properties}
\end{table}

@@

\begin{figure}[ht!]
	\centering
	\includegraphics[width=2.5in]{../Figures/HF_Center_TWNG_0.61m_Reps_Avg}
	\includegraphics[width=2.5in]{../Figures/HF_Center_TWNG_0.31m_Reps_Avg}\\
	\includegraphics[width=2.5in]{../Figures/HF_Center_TWNG_0.15m_Reps_Avg}
	\includegraphics[width=2.5in]{../Figures/HF_Center_TWNG_0m_Reps_Avg}\\
	\caption[Heat flux at the wall along the centerline of the natural gas fueled burner]{Heat flux at the wall along the centerline of the natural gas fueled burner. The heat fluxes are averages of two replicate experiments. Each graph displays data from a different distance between the edge of the burner and the instrumented wall. Beginning with the top left graph position and viewing left to right the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m respectively.}
	\label{HF_Center_TWNG_comp}
\end{figure}

\begin{figure}[ht!]
	\centering
	\includegraphics[width=2.5in]{../Figures/HF_Offset_TWNG_0.61m_Reps_Avg}
	\includegraphics[width=2.5in]{../Figures/HF_Offset_TWNG_0.31m_Reps_Avg}\\
	\includegraphics[width=2.5in]{../Figures/HF_Offset_TWNG_0.15m_Reps_Avg}
	\includegraphics[width=2.5in]{../Figures/HF_Offset_TWNG_0m_Reps_Avg}\\
	\caption[Heat flux at the wall along the edge of the natural gas fueled burner]{Heat flux at the wall along the edge of the natural gas fueled burner. The heat fluxes are averages of two replicate experiments. Each graph displays data from a different distance between the edge of the burner and the instrumented wall. Beginning with the top left graph position and viewing left to right the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m respectively.}
	\label{HF_Edge_TWNG_comp}
\end{figure}

@@

\Section{Summary}

These instrumented non-combustible wall experiments provide a data set for the fires conducted that can be compared against ignition data for gypsum wallboard from the literature to determine the potential for ignition of the paper surface of gypsum wallboard at a given distance from the wall with a given fuel. The Ignition Handbook by Babrauskas is the most comprehensive set of ignition data and related information ever written~\cite{Babrauskas:2003}.  Table 19 in the Ignition Handbook includes calculated ignition temperatures from experimental data from two separate source. Quintiere and Harkleroad conducted a study with the Lateral Ignition and Flame Spread (LIFT) apparatus, which they developed ~\cite{ASTM_E1321},\cite{Quitiere:1984}.  
The ignition temperature value calculated from the LIFT data for regular gypsum wallboard, 12.7~mm thick was 565 $\^circ$C with a critical heat flux for ignition of 35~kW/m^2.

The second source, was a series of cone calorimeter experiments conducted by Dillion~\cite{ASTM_E1354},~\cite{Dillion:1998}. The ignition temperature calculated from the cone calorimeter experiments for regular gypsum wallboard, 12.5~mm thick was 515 $\^circ$C with a minimum heat flux required for ignition of 26~kW/m^2.

A review of the graphs of the peak average temperatures at the wall, burner centerline array, shows that all of the fuels had at least one thermocouple position with a peak of approximately 500 $\^circ$C when the burner was against the wall.  At 0.15~m away from the wall the peak average temperature from this data set was provided by the gasoline fueled fire and it was approximately 350 $\^circ$C.  Based on temperature alone, the fires examined would only have the potential to ignite the paper surface of the gypsum wallboard when the burner was in contact with the wall.  The peak average temperatures at the wall, burner edge array, would indicate that the paper may not ignite at the burner edge.  The peak average temperature of approximately 400 $\^circ$C was generated by the gasoline fueled fire when the burner was positioned against the wall.

Of course the temperatures of the thermocouples in the wall as a result the of the energy transferred to the wall and the thermo-physical properties of the wall material. Assuming that the properties of calcium silicate board and the gypsum board are similar, let's review the heat flux graphs. All of the fuels generated heat fluxes in excess of 35~kW/m^2 at the wall, burner centerline array when the burners were against the wall.  The gasoline and the polyurethane fueled burners also exposed the wall to heat fluxes higher than 35~kW/^2 at 0.15~m from the wall.  However only the burner against the wall positions provided a combination of heat flux and temperature that had the potential to ignite the paper on the gypsum wallboard based on the handbook data.

@@

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|c|}
		\hline Density & \multicolumn{2}{c|}{Thermal Conductivity}   & \multicolumn{2}{c|}{Specific Heat}  \\
		\hline (kg/m$^3$) & Temperature ($^\circ$C) & (W/mK)  & Temperature ($^\circ$C)  & (kJ/kg $^\circ$C) \\
		\hline 807  & 24 	& 0.14 	&  		& 1.09 \\
		\hline  	& 100 	& 0.15  &  		&  \\
		\hline 790 	& 20 	& 0.18	& 20 	& 0.90 \\
		\hline  	& 100 	& 0.15 	& 100 	& 0.90 \\
		\hline  	& 200 	& 0.13 	& 200 	& 0.80 \\
		\hline  	& 500 	& 0.15 	& 500	& 0.90 \\
		\hline
		\end{tabular}
		\caption[Thermal properties for 12.7~mm gypsum wallboard]{Thermal properties for 12.7~mm gypsum wallboard}
		\label{tab:Gypsum wallboard_Thermal_Properties}
		\end{table}

@@

After receipt of the gypsum wallboard panels, the panels were laid out on the floor of the NIST Large Fire Research Laboratory, finished paper side up and painted with a roller.  The first coat was a latex primer, http://buyat.ppg.com/rep_pafpainttools_files/Pghpaints/tdb/78200.pdf.  Once the primer dried a top coat of white latex paint was applied http://buyat.ppg.com/rep_pafpainttools_files/Olympic/tdb/78010.pdf .  After the panels dried, they were weighed. The average mass of paint on the finished surface of the gypsum wallboard was 0.33~kg. Then the paint thickness was estimated an ultra sonic XXXX cite - in office.  Over the course of the experiments, random painted panels where checked with the ultra sonic device. The thickness of the paint varied from 0.05~mm to 0.13~mm.  Some of the painted panels were then cut-up into samples for the bench-scale experiments.

@@

The complete set of results from the instrumented gypsum wallboard experiments are provided in Appendix C.  Table ~\ref{tab:IWGB_experiments}    lists the experiments that were conducted for each fuel type.  In this chapter, as in the previous chapter the averaged peak values for temperatures and heat fluxes will presented at varying distances from the wall.  For the gasoline and polyurethane fires, an additional position, 0.08~m from the wall was added.

@@

\begin{sidewaystable}
	\centering
	\begin{tabular}{|c|c|c|c|c|c|}
	\hline
	Fuel (\# of tests)      &   Wall Type   &   Max. Height (m)	         & Max. Width (m)               &  Height at Max. Width (m) &  Area (m $^2$)  \\ \hline
		Natural Gas (10)    &   Open 		& 0.74 	$\pm$ 0.12 (16~\%)   &	0.24 $\pm$ 0.06 (25~\%) 	& 0.41 $\pm$ 0.07 (17~\%)	& 0.15 $\pm$ 0.05 (33~\%) 	\\
		Natural Gas (10)    &   Closed 		& 0.74	$\pm$ 0.06 (8~\%)    &	0.24 $\pm$ 0.02 (7~\%) 		& 0.43 $\pm$ 0.03 (7~\%)  	& 0.14 $\pm$ 0.01 (6~\%) 	\\
		Natural Gas (10)    &   Insulated 	& 0.71	$\pm$ 0.10 (14~\%)   &	0.23 $\pm$ 0.04 (17~\%) 	& 0.42 $\pm$ 0.07 (17~\%) 	& 0.15 $\pm$ 0.02 (14~\%)  	\\
		Gasoline (12)       &   Open    	& 0.83	$\pm$ 0.15 (18~\%)   &	0.28 $\pm$ 0.09 (32~\%) 	& 0.44 $\pm$ 0.14 (32~\%) 	& 0.17 $\pm$ 0.04 (24~\%)  	\\
		Gasoline (10)       &   Closed  	& 0.80	$\pm$ 0.10 (13~\%)   &	0.29 $\pm$ 0.04 (14~\%) 	& 0.44 $\pm$ 0.06 (14~\%) 	& 0.17 $\pm$ 0.04 (24~\%)  	\\
		Gasoline (10)       &   Insulated	& 0.82 	$\pm$ 0.09 (11~\%)   &	0.25 $\pm$ 0.08 (32~\%) 	& 0.40 $\pm$ 0.13 (32~\%) 	& 0.13 $\pm$ 0.04 (30~\%)  	\\
		\hline
	\end{tabular}
	\caption[Comparison of fire pattern dimensions as a function of wall construction type]{Comparison of fire pattern dimensions as a function of wall construction type. Values presented with 95~\% confidence limits.}
	\label{tab:Fire_Pattern_Dimensions_Wall_Construction}
\end{sidewaystable}

>>>>

A free standing steel frame wall with a 12.7~mm thick, 2.44~m high by 1.22~m wide, calcium silicate board face was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux. Marinite Type I is the commercial name of the non-combustible board used in these experiments.  It is normally used as a fire-resistant thermal insulation that also has structural strength. 90~\% of the board is composed of calcium silicate and calcium meta-silicate.  The remaining 10~\% of the board is composed of a combination of organic fibers, fiber glass filament and crystalline silica (quartz)~\cite{BNZ:MSDS}.  Material properties that impact heat transfer were selected from the manufacturer's literature and are presented in the table below~\cite{Marinite:1997}.

\begin{table}
	\centering
\begin{tabular}{|c|c|c|c|c|}
	\hline Density & \multicolumn{2}{c|}{Thermal Conductivity} & \multicolumn{2}{c|}{Specific Heat}   \\
    \hline (kg/m$^3$) & Temperature ($^\circ$C) & (W/mK)  & Temperature ($^\circ$C)  & (kJ/kg $^\circ$C) \\
	\hline 737  & 24 & 0.13 & 93 & 0.28 \\
	\hline  & 205 & 0.12  & 205 & 0.30 \\
	\hline  & 316 & 0.11 & 316 & 0.32 \\
	\hline  & 425 & 0.12 & 425 & 0.34 \\
	\hline  & 538 & 0.12 &  &  \\
	\hline
\end{tabular}
\caption[Thermal properties for Marinite I]{Thermal properties for Marinite I, as reported by the manufacturer BNZ3~\cite{Marinite:1997}}
\label{tab:Marinite_Thermal_Properties}
\end{table}

@@

\begin{figure}[ht!]
	\centering
	\includegraphics[width=2.5in]{../Figures/HF_Center_TWNG_0p61m_Reps_Avg}
	\includegraphics[width=2.5in]{../Figures/HF_Center_TWNG_0p31m_Reps_Avg}\\
	\includegraphics[width=2.5in]{../Figures/HF_Center_TWNG_0p15m_Reps_Avg}
	\includegraphics[width=2.5in]{../Figures/HF_Center_TWNG_0m_Reps_Avg}\\
	\caption[Heat flux at the wall along the centerline of the natural gas fueled burner]{Heat flux at the wall along the centerline of the natural gas fueled burner. The heat fluxes are averages of two replicate experiments. Each graph displays data from a different distance between the edge of the burner and the instrumented wall. Beginning with the top left graph position and viewing left to right the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m respectively.}
	\label{HF_Center_TWNG_comp}
\end{figure}

\begin{figure}[ht!]
	\centering
	\includegraphics[width=2.5in]{../Figures/HF_Offset_TWNG_0p61m_Reps_Avg}
	\includegraphics[width=2.5in]{../Figures/HF_Offset_TWNG_0p31m_Reps_Avg}\\
	\includegraphics[width=2.5in]{../Figures/HF_Offset_TWNG_0p15m_Reps_Avg}
	\includegraphics[width=2.5in]{../Figures/HF_Offset_TWNG_0m_Reps_Avg}\\
	\caption[Heat flux at the wall along the edge of the natural gas fueled burner]{Heat flux at the wall along the edge of the natural gas fueled burner. The heat fluxes are averages of two replicate experiments. Each graph displays data from a different distance between the edge of the burner and the instrumented wall. Beginning with the top left graph position and viewing left to right the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m respectively.}
	\label{HF_Edge_TWNG_comp}
\end{figure}

@@

\section{Summary}

These instrumented non-combustible wall experiments provide a data set for the fires conducted that can be compared against ignition data for gypsum wallboard from the literature to determine the potential for ignition of the paper surface of gypsum wallboard at a given distance from the wall with a given fuel. The Ignition Handbook by Babrauskas is the most comprehensive set of ignition data and related information ever written~\cite{Babrauskas:2003}.  Table 19 in the Ignition Handbook includes calculated ignition temperatures from experimental data from two separate source. Quintiere and Harkleroad conducted a study with the Lateral Ignition and Flame Spread (LIFT) apparatus, which they developed ~\cite{ASTM_E1321,Quintiere:1985}.
The ignition temperature value calculated from the LIFT data for regular gypsum wallboard, 12.7~mm thick was 565~$^{\circ}$C with a critical heat flux for ignition of 35~kW/m$^2$.

The second source, was a series of cone calorimeter experiments conducted by Dillon~\cite{ASTM_E1354,Dillon:1998}. The ignition temperature calculated from the cone calorimeter experiments for regular gypsum wallboard, 12.5~mm thick was 515~$^{\circ}$C with a minimum heat flux required for ignition of 26~kW/m$^2$.

A review of the graphs of the peak average temperatures at the wall, burner centerline array, shows that all of the fuels had at least one thermocouple position with a peak of approximately 500~$^{\circ}$C when the burner was against the wall.  At 0.15~m away from the wall the peak average temperature from this data set was provided by the gasoline fueled fire and it was approximately 350~$^{\circ}$C.  Based on temperature alone, the fires examined would only have the potential to ignite the paper surface of the gypsum wallboard when the burner was in contact with the wall.  The peak average temperatures at the wall, burner edge array, would indicate that the paper may not ignite at the burner edge.  The peak average temperature of approximately 400~$^{\circ}$C was generated by the gasoline fueled fire when the burner was positioned against the wall.

Of course the temperatures of the thermocouples in the wall as a result the of the energy transferred to the wall and the thermo-physical properties of the wall material. Assuming that the properties of calcium silicate board and the gypsum board are similar, let's review the heat flux graphs. All of the fuels generated heat fluxes in excess of 35~kW/m$^2$ at the wall, burner centerline array when the burners were against the wall.  The gasoline and the polyurethane fueled burners also exposed the wall to heat fluxes higher than 35~kW/$^2$ at 0.15~m from the wall.  However only the burner against the wall positions provided a combination of heat flux and temperature that had the potential to ignite the paper on the gypsum wallboard based on the handbook data.

@@

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|c|}
		\hline Density & \multicolumn{2}{c|}{Thermal Conductivity}   & \multicolumn{2}{c|}{Specific Heat}  \\
		\hline (kg/m$^3$) & Temperature ($^{\circ}$C) & (W/mK)  & Temperature ($^{\circ}$C)  & (kJ/kg $^{\circ}$C) \\
		\hline 807  & 24 	& 0.14 	&  		& 1.09 \\
		\hline  	& 100 	& 0.15  &  		&  \\
		\hline 790 	& 20 	& 0.18	& 20 	& 0.90 \\
		\hline  	& 100 	& 0.15 	& 100 	& 0.90 \\
		\hline  	& 200 	& 0.13 	& 200 	& 0.80 \\
		\hline  	& 500 	& 0.15 	& 500	& 0.90 \\
		\hline
		\end{tabular}
		\caption[Thermal properties for 12.7~mm gypsum wallboard]{Thermal properties for 12.7~mm gypsum wallboard}
		\label{tab:Gypsum wallboard_Thermal_Properties}
		\end{table}

@@

After receipt of the gypsum wallboard panels, the panels were laid out on the floor of the NIST Large Fire Research Laboratory, finished paper side up and painted with a roller.  The first coat was a latex primer, http://buyat.ppg.com/rep\_pafpainttools\_files/Pghpaints/tdb/78200.pdf.  Once the primer dried a top coat of white latex paint was applied http://buyat.ppg.com/rep\_pafpainttools\_files/Olympic/tdb/78010.pdf.  After the panels dried, they were weighed. The average mass of paint on the finished surface of the gypsum wallboard was 0.33~kg. Then the paint thickness was estimated an ultra sonic XXXX cite - in office.  Over the course of the experiments, random painted panels where checked with the ultra sonic device. The thickness of the paint varied from 0.05~mm to 0.13~mm.  Some of the painted panels were then cut-up into samples for the bench-scale experiments.

@@

The complete set of results from the instrumented gypsum wallboard experiments are provided in Appendix C. Table~\ref{tab:IWGB_experiments} lists the experiments that were conducted for each fuel type.  In this chapter, as in the previous chapter the averaged peak values for temperatures and heat fluxes will presented at varying distances from the wall.  For the gasoline and polyurethane fires, an additional position, 0.08~m from the wall was added.

@@

\begin{sidewaystable}
	\centering
	\begin{tabular}{|c|c|c|c|c|c|}
	\hline
	Fuel (\# of tests)      &   Wall Type   &   Max. Height (m)	         & Max. Width (m)               &  Height at Max. Width (m) &  Area (m$^2$)  \\ \hline
		Natural Gas (10)    &   Open 		& 0.74 	$\pm$ 0.12 (16~\%)   &	0.24 $\pm$ 0.06 (25~\%) 	& 0.41 $\pm$ 0.07 (17~\%)	& 0.15 $\pm$ 0.05 (33~\%) 	\\
		Natural Gas (10)    &   Closed 		& 0.74	$\pm$ 0.06 (8~\%)    &	0.24 $\pm$ 0.02 (7~\%) 		& 0.43 $\pm$ 0.03 (7~\%)  	& 0.14 $\pm$ 0.01 (6~\%) 	\\
		Natural Gas (10)    &   Insulated 	& 0.71	$\pm$ 0.10 (14~\%)   &	0.23 $\pm$ 0.04 (17~\%) 	& 0.42 $\pm$ 0.07 (17~\%) 	& 0.15 $\pm$ 0.02 (14~\%)  	\\
		Gasoline (12)       &   Open    	& 0.83	$\pm$ 0.15 (18~\%)   &	0.28 $\pm$ 0.09 (32~\%) 	& 0.44 $\pm$ 0.14 (32~\%) 	& 0.17 $\pm$ 0.04 (24~\%)  	\\
		Gasoline (10)       &   Closed  	& 0.80	$\pm$ 0.10 (13~\%)   &	0.29 $\pm$ 0.04 (14~\%) 	& 0.44 $\pm$ 0.06 (14~\%) 	& 0.17 $\pm$ 0.04 (24~\%)  	\\
		Gasoline (10)       &   Insulated	& 0.82 	$\pm$ 0.09 (11~\%)   &	0.25 $\pm$ 0.08 (32~\%) 	& 0.40 $\pm$ 0.13 (32~\%) 	& 0.13 $\pm$ 0.04 (30~\%)  	\\
		\hline
	\end{tabular}
	\caption[Comparison of fire pattern dimensions as a function of wall construction type]{Comparison of fire pattern dimensions as a function of wall construction type. Values presented with 95~\% confidence limits.}
	\label{tab:Fire_Pattern_Dimensions_Wall_Construction}
\end{sidewaystable}
